OptiFit: a fast method for fitting amplicon sequences to existing OTUs

### 2021-02-23 paper/paper.Rmd
Problem statement

There are three main categories of OTU clustering algorithms: closed reference,
open reference, and _de novo_ clustering. Closed reference methods assign
sequences to a set of pre-made OTUs generated from reference sequences. If a
query sequence is not within the distance threshold to any of the reference
sequences, it is discarded. While reference-based clustering is generally fast,
it is limited by the diversity represented in the reference database. _De novo_
methods cluster sequences based on their distance to each other, without the use
of an external reference. open reference methods first perform closed reference
clustering, then any sequences that cannot be assigned to reference OTUs are
clustered _de novo_ to create additional OTUs.  _De novo_ clustering overcomes
the limitations of reference databases by considering only sequences in the
dataset, but is more computationally intensive and generates different OTU
assignments when new sequences are introduced. Previous studies found that _de
novo_ clustering created the highest quality OTU assignments based on the
Matthews correlation coefficient (MCC) [@schloss_application_2016;
@westcott_novo_2015].

- **TODO: clustering methods. describe reference vs de novo clustering. trade-offs. machine learning use-case.**
- **TODO: mention OptiClust and vsearch.**
- **TODO:** gap in knowledge/ problem statement. acknowledge knock against OTUs, that they're not reproducible across studies. idea of ref-clustering is you can use existing OTUs to compare data across studies. can we make a ref-based method that's as good as _de novo_.
- **TODO:** in this study, we fix that problem.

>>>>

There are three main categories of OTU clustering algorithms: closed reference,
open reference, and _de novo_ clustering. Closed reference methods assign
sequences to a set of pre-made OTUs generated from reference sequences. If a
query sequence is not within the distance threshold to any of the reference
sequences, it is discarded. While reference-based clustering is generally fast,
it is limited by the diversity represented in the reference database. Rare or
novel sequences in the sample will be lost if they are not represented in the
database. _De novo_ methods cluster sequences based on their distance to each
other, without the use of an external reference ._De novo_ clustering overcomes
the limitations of reference databases by considering only sequences in the
dataset, but is more computationally intensive and generates different OTU
assignments when new sequences are introduced.  Open reference methods take a
hybrid approach, first performing closed reference clustering, then any
sequences that cannot be assigned to reference OTUs are clustered _de novo_ to
create additional OTUs. Previous studies found that the OptiClust _de novo_
clustering algorithm created the highest quality OTU assignments based on the
Matthews correlation coefficient (MCC) [@schloss_application_2016;
@westcott_novo_2015]. However, a disadvantage of _de novo_ clustering is that
different input samples will result in different OTUs, making it difficult to
compare taxonomic composition of communities between studies or to use machine
learning models trained with _de novo_ OTUs to make predictions on new data.

To overcome the limitations of _de novo_ clustering while maintaining OTU
quality, we developed OptiFit, a reference-based clustering algorithm in the
mothur program which takes existing OTUs as the reference to fit new sequences
to. Here, we tested the OptiFit algorithm with databases or _de novo_ OTUs as
the reference and compared the performance to existing tools.
